---
title: "Practical Machine Learning Course Project"
author: "Jeffrey M. Hunter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    keep_md: no
    df_print: paged
    css: css/custom.css
  pdf_document:
    toc: yes
    df_print: kable
    number_sections: false
    fig_caption: yes
    highlight: tango
    dev: pdf
  word_document:
    toc: yes
    df_print: paged
    keep_md: no
---

## Synopsis

This report contains research that was conducted for the Practical Machine
Learning course as part of the Data Science Specialization offered through
Coursera from Johns Hopkins University. This writeup was built in RStudio
using the knitr function to publish the final report in HTML format.

The source code for this project can be found on
GitHub: <a target="_blank" href="https://github.com/oraclejavanet/practical-machine-learning-course-project">Practical Machine Learning Course Project</a>

## Introduction

Using devices such as *Jawbone Up*, *Nike FuelBand*, and *Fitbit* it is now
possible to collect a large amount of data about personal activity relatively
inexpensively. These type of devices are part of the quantified self movement –
a group of enthusiasts who take measurements about themselves regularly to
improve their health, to find patterns in their behavior, or because they are
tech geeks. One thing that people regularly do is quantify how much of a
particular activity they do, but they rarely quantify how well they do it.

The goal for this assignment is to analyze the *Weight Lifting Exercise*
dataset and develop a machine learning algorithm using biometric data to predict
the manner in which 6 participants performed a particular dumbbell exercise.
This is the "classe" variable in the training set.

The biometric data was collected from accelerometers on the belt, forearm, arm,
and dumbbell of the 6 participants. The participants were asked to perform
barbell lifts correctly and incorrectly in 5 different ways.

> For more information about the dataset used for this analysis, see section
<a target="_blank" href="http://groupware.les.inf.puc-rio.br/har#weight_lifting_exercises">Weight Lifting Exercises Dataset</a> from the Human Activity Recognition (HAR) project website.

The resulting prediction model will also serve as the basis for the
"Course Project Prediction Quiz Portion" of this assignment where the machine
learning algorithm will be applied to 20 test cases available in the test data.
The predictions will submitted in the appropriate format for automated grading.

## Environment Setup

Prepare the session by loading all necessary packages and clearing the global
workspace (including hidden objects).

```{r load-packages, message = FALSE, echo = TRUE}
library(knitr)
library(ggplot2)
library(caret)
library(randomForest)
rm(list = ls(all.names = TRUE))
setwd("~/repos/coursera/github-assignments/practical-machine-learning-course-project")
```

```{r setup, include = FALSE}
# set knitr options
knitr::opts_chunk$set(echo = TRUE, fig.path = 'figures/')

# free up memory and display statistics on free memory
gc()

# disable scientific notation for numbers
options(scipen = 1)

# knit hook to allow partial output from a code chunk
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options)) # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) { # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```

## Prepare the Data

### Load the Data

```{r load-data, echo = TRUE}
trainURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
validationURL  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

trainDataFile <- "data/pml-training.csv"
validationDataFile <- "data/pml-testing.csv"

if (!file.exists('data')) {
    dir.create('data')
}
if (!file.exists(trainDataFile)) {
    download.file(url = trainURL, destfile = trainDataFile)
}
if (!file.exists(validationDataFile)) {
    download.file(url = validationURL, destfile = validationDataFile)
}

trainData <- read.csv(trainDataFile, sep = ",", header = TRUE)
validationData <- read.csv(validationDataFile, sep = ",", header = TRUE)

stopifnot(file.size(trainDataFile) == 12202745)
stopifnot(file.size(validationDataFile) == 15113)
```

```{r dataset-dimensions, echo = TRUE}
dim(trainData)
dim(validationData)
```

The training dataset contains `r dim(trainData)[1]` observations and
`r dim(trainData)[2]` variables while the testing dataset (used for the
prediction quiz portion of the assignment) contains `r dim(validationData)[1]`
observations and `r dim(validationData)[2]` variables.

### Clean the Data

*Pending*

```{r clean-data, echo = TRUE}
```

### Partition the Data

In order to accurately estimate the out-of-sample error, partition the training
dataset (`trainData`) into two sets: 60% of the training data for the modeling
process and the remaining 40% for the test set. This will be performed using
random subsampling without replacement.

```{r partition-data, echo = TRUE}
inTrain  <- createDataPartition(trainData$classe, p = 0.6, list = FALSE)
trainSet <- trainData[inTrain, ]
testSet  <- trainData[-inTrain, ]
```

The `validationData` test dataset will only be used for the "Course Project
Prediction Quiz Portion" of this assignment and thus will remain unchanged.

### Exploratory Data Analysis

The 6 young and healthy participants were asked to perform one set of 10
repetitions of the *Unilateral Dumbbell Biceps Curl* in the following 5
different fashions:

* Class A: exactly according to the specification
* Class B: throwing the elbows to the front
* Class C: lifting the dumbbell only halfway
* Class D: lowering the dumbbell only halfway
* Class E: throwing the hips to the front

This is the "classe" variable in the training set and is used as an indicator
of the training outcome. Class A corresponds to the specified execution of the
exercise, while the other 4 classes correspond to common mistakes.

The following plot shows the distribution of the classe variable in the training
set.

```{r dist-classe-var-training-set, echo = TRUE}
g <- ggplot(trainSet, aes(classe))
g <- g + geom_bar(fill = rgb(0.2, 0.4, 0.6, 0.8))
g <- g + xlab("Classe Level")
g <- g + ylab("Frequency")
g <- g + theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(hjust = 0.5, vjust = 0.5),
               axis.text.y = element_text(hjust = 0.5, vjust = 0.5))
g <- g + ggtitle("Distribution of the Classe Variable in the Training Set")
print(g)
```

## Prediction Model Building

*Pending*

### Random Forest

*Pending*

### Decision Tree

*Pending*




